### 2022-02-09 article.tex
We Ain't Need No Dependent Types!

\subsection{A Lightweight Form of Dependent Typing}
Match types enable a lightweight form of dependent typing, since term-level pattern matching expressions can be typed accordingly at the type level as a match types.
Consider the following function definition:
%
\begin{lstlisting}
def elem[X](x: X): Elem[X] = x match {
  case x: String => x.charAt(0)
  case x: List[t] => elem(x.head)
  case x: Any => x
}
\end{lstlisting}
%
This definition is well-typed because the match expression in |elem|'s body has the exact same scrutinee and pattern types as |Elem[X]| (the function's return type).

>>>>

\subsection{A Lightweight Form of Dependent Typing}
\label{subsec:a-lightweight-form-of-dependent-typing}

Match types enable a lightweight form of dependent typing, since term-level pattern matching expressions can be typed accordingly at the type level as a match types.
Consider the following function definition:
%
\begin{lstlisting}
def elem[X](x: X): Elem[X] = x match {
  case x: String => x.charAt(0)
  case x: List[t] => elem(x.head)
  case x: Any => x
}
\end{lstlisting}
%
This definition is well-typed because the match expression in |elem|'s body has the exact same scrutinee and pattern types as |Elem[X]| (the function's return type).

### 2022-02-09 symposium.tex
We Ain't Need No Dependent Types!

  \item The term-level capturing group processing, which extracts and sanitizes the output of the regex engine, in accordance with the representation computed at the type level (\Cref{sec:term-level}).

@@

\regexUserLevel
% object Regex:
%   def apply[R <: String & Singleton](regex: R): Regex[Compile[R]] =
%     new Regex(regex)
%
% class Regex[P] private (val regex: String):
%   val pattern = Pattern.compile(regex)
%   def unapply(s: String): Option[P] =
%     val m = pattern.matcher(s)
%     if (m.matches())
%       val a = Array.tabulate(m.groupCount) { i => m.group(i + 1) }
%       Some(toTuple[P](regex, a))
%     else
%       None

@@

The main difference between Scala's original |Regex| implementation and ours lies in the signature of the |unapply| method, which determines the nature of extraction performed by that method.
By returning an optional tuple, we tell the compiler that a successful matching consists of exactly $n$ capturing groups, where $n$ corresponds to the tuple arity, and that the type of those groups will correspond to the tuple types.
We take advantage of that flexibility to enrich the results of Java's regex engine by wrapping nullable values into options.
In our implementation, this is done via the |toTuple| method, which is our
term-level capturing group processing component.
As a result, our implementation never returns null values, which removes the null-checking burden from the user.

@@

\subsection{Non-Capturing Groups}

@@

We handle nested groups by updated our algorithm to operate in two different modes while iterating through the regex's characters.
The algorithm can either be outside of an optional group (|Opt=0|), in which case it computes nullability of newly encountered via |IsNullable|, or inside an optional group (|Opt>0|), in which case it treats every group as nullable.
This last improvement leads to the final iteration of our type-level program handles nested groups by operating in two different modes.

With this latest improvement, we develop our implementation into its final iteration (|IsCapturing| omitted for brevity):

@@

>>>>

  \item The term-level capturing group processing, which extracts and sanitizes the output of the regex engine, in accordance with the previously computed type-level representation computed (\Cref{sec:term-level}).

@@

\regexUserLevel
% object Regex:
%   def apply[R <: String & Singleton](regex: R): Regex[Compile[R]] =
%     new Regex(regex)
%
% class Regex[P] private (val regex: String):
%   val pattern = Pattern.compile(regex)
%   def unapply(s: String): Option[P] =
%     val m = pattern.matcher(s)
%     if (m.matches())
%       val a = Array.tabulate(m.groupCount)(i => m.group(i + 1))
%       Some(transform[P](regex, a))
%     else
%       None

@@

The main difference between Scala's original |Regex| implementation and ours lies in the signature of the |unapply| method, which determines the nature of extraction performed by that method.
By returning an optional tuple, we tell the compiler that a successful matching consists of exactly $n$ capturing groups, where $n$ corresponds to the tuple arity, and that the type of those groups will correspond to the tuple types.
We take advantage of that flexibility to enrich the results of Java's regex engine by wrapping nullable values into options.
In our implementation, this is done via the |transform| method, which is our
term-level processing component.
As a result, our implementation never returns null values, which removes the null-checking burden from the user.

@@

\subsection{Non-Capturing Groups}
\label{subsec:non-capturing-groups}

@@

We handle nested groups by updated our algorithm to operate in two different modes while iterating through the regex's characters.
Our algorithm can either be outside of an optional group (|Opt=0|), in which case it computes nullability of newly encountered groups via |IsNullable|, or inside an optional group (|Opt>0|), in which case it treats every group as nullable.
This last improvement leads to the final iteration of our type-level program handles nested groups by operating in two different modes.

With this latest improvement, we advance our implementation to its final iteration (we omit usages of |IsCapturing|, from \Cref{subsec:non-capturing-groups}, for brevity):

@@

The term-level component of our library is in charge of sanitizing and packaging the results of regular expression matchings.
Concretely, this component's job consists in transforming the string array that comes out of the regex engine into an appropriately sized tuple, with nullable elements wrapped in options.
That transformation must conform to the representation computed at the type level.

In Scala, all type parameters are erased, which can lead so some friction when trying to write programs whose execution \emph{depends} on a type, such as in the problem at hand.
In this section, we propose two different solutions to that problem.

\begin{enumerate}
  \item In \Cref{subsec:we-ain-t-need-no-dependent-types}, we proceed by sheer force of code duplication: we translate the entirety of our type-level algorithm into term-level functions and use a cast to correlate the two.

  \item In \Cref{subsec:implicit-based-extractor-synthesis}, we show...
\end{enumerate}

\subsection{We Ain't Need No Dependent Types!}
\label{subsec:we-ain-t-need-no-dependent-types}

The first implementation of our term-level component duplicates the type-level definitions presented in \Cref{sec:type-level} to compute a list of sanitizing functions that correspond to the given regex's capturing groups.
In a nutshell, instead of accumulating |String| and |Option[String]| types, we rewrite the outer loop of our algorithm to accumulate identity functions and eta-expansions of option's constructor:

\regexTermLvlLoop
% val identity: String => Any = { x => assert(x != null); x }
% type Accumulator = List[String => Any]
%
% def loop(r: String, lo: Int, hi: Int, acc: Accumulator, opt: Int): Accumulator =
%   lo match
%     case `hi` => acc
%     case _  => r.charAt(lo) match
%       case ')' => loop(r, lo + 1, hi, acc, 0.max(opt - 1))
%       case '(' =>
%         opt match
%           case 0 => isNullable(r, lo + 1, hi, 0) match
%             case true  => loop(r, lo + 1, hi, Option.apply :: acc, 1)
%             case false => loop(r, lo + 1, hi, identity :: acc, 0)
%           case _ => loop(r, lo + 1, hi, Option.apply :: acc, opt + 1)
%       case '\\' => loop(r, lo + 2, hi, acc, opt)
%       case _ => loop(r, lo + 1, hi, acc, opt)

\noindent
After having computed those functions, we simply do a pointwise application with the raw output of the regex engine, package the result into a tuple, \emph{et voilà!}

\regexTransform
% def transform[P](pattern: String, arr: Array[String]): P =
%   val fs = loop(pattern, 0, pattern.length, Nil, 0).reverse
%   val ts = Tuple.fromArray(arr.zip(fs).map { (x, f) => f(x) })
%   assert(arr.size == fs.size, "Unexpected number of capturing groups")
%   ts.asInstanceOf[P]

\emph{In principle}, Scala~3's type system is capable of doing a lightweight form of dependent typing (\Cref{subsec:a-lightweight-form-of-dependent-typing}), where it...

\begin{lstlisting}
def isMarked
  [R <: String, At <: Int, Hi <: Int]
  (r: R, at: At, hi: Hi)
  : IsMarked[R, At, Hi]
\end{lstlisting}

\subsection{Implicit-Based Extractor Synthesis}
\label{subsec:implicit-based-extractor-synthesis}
